---
title: "LLM coding agent"
format: html
editor: visual
---

# Overview

This brief tutorial provides a hands on demonstration of a simple coding agent based on Hadley Wickham's recent blog post

<https://tidydesign.substack.com/p/a-coding-agent-is-six-functions-in>

The ai_lecture discussed how LLMs are just next word generators.  This tutorial demonstrates the development of a harness for coding.


## Set up

First install the latest version of the package "ellmer".  Ellmer is a tidyverse associated package for interfacing with LLMs through R.

```{r}
#| eval: false
remotes::install_github("tidyverse/ellmer")
```

Second, make sure that your API key is in the creds.R file (see README.md for more information)

```{r}
source("creds.R")
```

Third, load the ellmer package

```{r}
library(ellmer)
```

## Example 1

*Goal: see that an LLM can converse but cannot act on its own — it has no way to write files or run code.*

Here is an example of ellmer usage.  The command below start a chat with the `MiniMax-M2.7" model on VT's LLM service.  Per VT: "Use gpt-oss-120b for fast responses, GLM-5.2 or Kimi-K2.6 for complex tasks, and MiniMax-M2.7 for software engineering."

```{r}
chat <- ellmer::chat_openai_compatible(
  model = "MiniMax-M2.7",
  base_url = "https://llm-api.arc.vt.edu/api/v1/"
)
```

Now we can chat with the LLM.  You should see text describing the difference between a tibble and a data frame but it didn't save a file because it doesn't know how.

```{r}
chat$chat("What is the difference between a tibble and a data frame?  Save the output as a markdown file")
```

## Example 2

*Goal: give the LLM tools for reading, writing, and running files, and watch it complete a task the first example couldn't.*

Now we are going to add a few tools to the harness.  The tools are reading files from your computer, writing files from your computer, and running commands on your computer.  Running commands on your computer is the critical tool for running and evaluating code.  

```{r}
read_file <- function(path) {
  paste(readLines(path), collapse = "\n")
}

write_file <- function(path, content) {
  writeLines(content, path)
  paste0("Wrote ", path, ".")
}

run_command <- function(command) {
  paste(system(command, intern = TRUE), collapse = "\n")
}
```

We are going to start a new chat but provide a bit more context through the `system_prompt`.


```{r}
chat <- ellmer::chat_openai_compatible(
  model = "MiniMax-M2.7",
  base_url = "https://llm-api.arc.vt.edu/api/v1/",
  system_prompt = "
    You are a coding assistant working in the user's current directory.
    Use the tools to write files to the current directory.
  "
)
```

Now we need to register the tools with the LLM.  Registering involves providing the function, a description of what the function does so it knows when to use it, and arguments required by the function.

```{r}
chat$register_tool(tool(
  read_file,
  description = "Read the entire contents of a text file.",
  arguments = list(
    path = type_string("Path to the file, relative to the working directory.")
  )
))

chat$register_tool(tool(
  write_file,
  description = "Write content to a file, overwriting it if it already exists.
    Always read the file first if you only want to change part of it.",
  arguments = list(
    path = type_string("Path to the file, relative to the working directory."),
    content = type_string("The full new contents of the file.")
  )
))

chat$register_tool(tool(
  run_command,
  description = "Run a shell command and return its output. Use this to list
    files (ls), search code (grep), run tests, or use git.",
  arguments = list(
    command = type_string("The shell command to run.")
  )
))
```

Now we can do the same chat but with the tooling harness included

```{r}
chat$chat("What is the difference between a tibble and a data frame?  Save the output as a markdown file")
```

Look in your working directory — there should be a new .md file.  The LLM wrote the file to your computer!

> **Note on safety:** `run_command` can execute any shell command, which is powerful but risky. Running the tutorial inside Docker (as described in the README) limits what the agent can reach — it can only touch files inside the container.

## Example 3

*Goal: upgrade the system prompt so the agent loops autonomously until the task is complete, and add safety helpers that prevent it from escaping the project directory.*

Example three shows how the harness can be used to write code from scratch that accomplishes a task.

First we need to expand our `system_prompt` so that the LLM will work until it has finished the task.

```{r}
chat <- ellmer::chat_openai_compatible(
  model = "MiniMax-M2.7",
  base_url = "https://llm-api.arc.vt.edu/api/v1/",
  system_prompt = "
    You are a coding assistant working in the user's current directory.
    Use the tools to explore and modify their project. Before editing a
    file, read it first. After making changes, run any relevant tests or
    checks to confirm your work. Keep going until the task is done.  You prefer to 
    write code in R.
  "
)
```

We need to add some new functions.  You can read more about them in Hadley's post.  They are used to provide some safety so the agent only works in your working directory.

```{r}
list_files <- function(path = ".") {
  paste(list.files(path), collapse = "\n")
}

safe_path <- function(path) {
  root <- normalizePath(".", winslash = "/", mustWork = TRUE)
  full <- normalizePath(path, winslash = "/", mustWork = FALSE)
  within_root <- full == root || startsWith(full, paste0(root, "/"))
  
  if (!within_root) {
    stop("Path is outside the project directory: ", path)
  }
  path
}

search_files <- function(pattern, path = ".") {
  files <- list.files(safe_path(path), recursive = TRUE, full.names = TRUE)
  
  hits <- character()
  for (file in files) {
    lines <- readLines(file, warn = FALSE)
    matches <- grep(pattern, lines)
    hits <- c(hits, paste0(file, ":", matches, ": ", lines[matches]))
  }
  paste(hits, collapse = "\n")
}
```

Now register these new tools along with our previous tools.  You have to do this every time you start a new chat with `chat <- ellmer::chat_openai_compatible` (note: you can use `chat$set_turns(list())` to reset a chat and keep the tools, but we restarted the chat from scratch because we edited the system prompt.).  

Notice how none of our tools are specific to any code problem.  They are generic tools for reading, writing, and executing files.

```{r}
chat$register_tool(tool(
  read_file,
  description = "Read the entire contents of a text file.",
  arguments = list(
    path = type_string("Path to the file, relative to the working directory.")
  )
))

chat$register_tool(tool(
  write_file,
  description = "Write content to a file, overwriting it if it already exists.
    Always read the file first if you only want to change part of it.",
  arguments = list(
    path = type_string("Path to the file, relative to the working directory."),
    content = type_string("The full new contents of the file.")
  )
))

chat$register_tool(tool(
  run_command,
  description = "Run a shell command and return its output. Use this to list
    files (ls), search code (grep), run tests, or use git.",
  arguments = list(
    command = type_string("The shell command to run.")
  )
))


chat$register_tool(tool(
  list_files,
  description = "List the files and directories at a given path.",
  arguments = list(
    path = type_string("Directory to list. Defaults to the working directory.")
  )
))

chat$register_tool(tool(
  search_files,
  description = "Search the contents of all files for a regular expression,
    returning matching lines as 'path:line: text'.",
  arguments = list(
    pattern = type_string("A regular expression to search for."),
    path = type_string("Directory to search. Defaults to the working directory.")
  )
))
```

The prompt below asks the agent to find and use the Mauna Loa CO₂ data that is included in this repo, so the agent should discover it via `run_command`.

```{r}
chat$chat("Find the Mauna Loa data.  Create an R script to plot the data using ggplot. Save the plot as a png.")
```

Confirm that an R script and plot were created in your working directory.

Now we can continue the chat to revise the plot

```{r}
chat$chat("A horizonal and vertical line for the first year that the curve was always above 400 ppm")
```

This update to the chat included through previous chat and added the additional request.  You can see the full chat by running the following.  It will produce a lot of text (you can click on the x on the right side below the gray code chuck to remove the text after you see what it is doing).  The key thing to understand is that all of the text below is passed to the LLM every time you chat so long chat requests, multiple chat calls, and long responses from LLM request in a long message.  The length of this message is approximately your "context window" for the LLM and is related to the number of tokens it uses.  LLM performance degrades when the context window gets too long.

```{r}
#| eval: false
chat
```

## Example 4

*Goal: try the agent yourself with your own prompt.*

If you want to test the coding agent out for yourself, you can reset the chat but keep the tools by running the following command

```{r}
#| eval: false
chat$set_turns(list())
```

```{r}
#| eval: false
chat$chat("ADD YOUR CHAT MESSAGE HERE")
```

## Take home

Enjoy and remember that Claude Code is just version of these that has more extensive system prompts and uses the Claude LLM models!

For example, there have been attempts to reverse engineer Claude Code.  See this repository with many instructions as part of the system prompt. (https://github.com/Piebald-AI/claude-code-system-prompts)


